First Season MWA EoR Power Spectrum Results at Redshift 7

### 2016-05-26 ms.tex
Add files via upload
Updated ms.tex with Cath's comments and some text for the ref pipeline section. All changes/suggestions made with \textbf{}.

Using the first generation of instruments, several upper limits have been placed on the 21 
cm EoR signal \citep{Ali:2015, Dillon:2014, Parsons:2014, Jacobs:2015, Paciga:2013}, and 
\citealt{Pober:2015} was able to place constraints on physical reionization models. 
However, much work is to be done to understand the data produced by these arrays. 
Foreground subtraction and isolation has emerged as a priority in the field. Central to most 
analysis strategies is the concept of an ``EoR window" -- the region of Fourier space where 
spectrally smooth foregrounds have been isolated from the isotropic (spherically symmetric 
in Fourier space) cosmological signal \citep{Morales:2006, Bowman:2009}. More recent 
studies have shown the existence of a foreground ``wedge", the result of instrumental mode 
mixing, throwing power from spectrally smooth foregrounds to higher Fourier modes 
\citep{Thyagarajan:2015b, Thyagarajan:2015, Trott:2014, Liu:2014a, Liu:2014b, 
Hazelton:2013, Pober:2013, Vedantham:2012, Morales:2012, Datta:2010}. Despite this 
setback, the EoR window is preserved above the wedge, and the field is pushing forward with 
several analysis pipelines under active development \citep[e.g. B.~J.~Hazelton et al. 2016, in 
prep; D.~A.~Mitchell et al. 2016, in prep; D.~C.~Jacobs et al. 2016, in review; ][]{Trott:2016, 
Dillon:2013, Trott:2012}.

@@

The remainder of this paper is organized as follows. In Section~\ref{sec:MWA} we briefly 
describe the MWA instrument and the observations used in this work, in Section~
\ref{sec:analysis} we describe our analysis pipeline, in Section~\ref{sec:techniques} we 
describe several novel techniques in our analysis, in Section~\ref{sec:deep} we discuss our 
efforts to apply the analysis pipeline to a deep data integration, and we discuss future work 
in Section~\ref{sec:discussion}. Throughout this paper we use a $\Lambda$CDM 
cosmology with $\Omega_m=0.73$, $\Omega_\Lambda=0.27$, and $h = 0.7$, consistent 
with WMAP seven year results \citep{Komatsu:2011}. All distances and wavenumbers are in 
comoving coordinates.

@@

The technical design of the MWA is reviewed in \citealt{Tingay:2013}, and we highlight a 
few key characteristics that will become important in our analysis of the data. Each 
antenna of the MWA comprises 16 dual-polarization dipoles placed on a regular grid, lying 
on a ground screen. The radio frequency signals from these dipoles feed into an analog 
beamformer which uses physical delays to ``point" the antenna. The MWA contains 128 of 
these antennas, with a tightly packed core of radius 50 m, and extending out to a radius 
1.5 km for higher resolution calibration and imaging.

@@

\section{Analysis Pipeline}\label{sec:analysis}
In order to ensure consistency in analysis, the international MWA EoR collaboration has 
defined two independent reference analysis pipelines. The details of this strategy are 
outlined in D.~C.~Jacobs et al. 2016 (in review). This study is based on the FHD-to-
\eppsilon pipe, which is in turn based on the Fast Holographic Deconvolution (FHD
\footnote{\url{https://github.com/EoRImaging/FHD}}, \citealt{Sullivan:2012}) and Error 
Propagated Power Spectrum with Interleaved Observed Noise (\eppsilon
\footnote{\url{https://github.com/EoRImaging/eppsilon}}, B.~J.~Hazelton et al. 2016, in 
prep) packages. The general flow of the pipeline is shown in Figure~\ref{fig:pipe}, and we 
give an overview below.

@@

Besides RFI, we manually flag data to remove three other effects. As mentioned in 
Section~\ref{sec:MWA}, the edges of the coarse bands contain high levels of aliasing from 
the filter applied in the digital receivers. We flag two 40 kHz channels on either side of the 
coarse band edges (total of four channels per coarse band edge). In principle the aliasing 
could be calibrated out and these channels could be recovered, but this is left for future 
work. Second, early data from the MWA also contained some instances where antennas 
were not pointed properly at the beginning of observations due to a beamformer error. This 
problem has since been resolved, but exists in our data. We therefore flag the first two 
seconds of each observation to avoid any potentially mis-pointed data. Finally, we flag the 
central fine channel of each coarse band because it corresponds to the DC mode of that 
sub-band and can easily float, washing out useful information.

@@

\section{Novel Analysis Methods}\label{sec:techniques}
\textcolor{red}{Needs a better section title}

The two dimensional power spectrum is a useful figure of merit (FoM) as we improve and 
refine our analysis pipeline. Foregrounds and systematics often manifest with characteristic 
shapes in this space, enabling us to diagnose problems and quantify improvements 
\citep{Morales:2012}. We use the MWA 3 hour ``golden'' data set from August 23, 2013, to 
repeatedly form power spectra to test and refine our analysis. While an exhaustive catalog 
of these improvements would be unsuited for this article (and likely bore the reader), we 
demonstrate the utility of the 2D power spectrum as a FoM with a few key techniques we 
have employed in our analysis.

@@

Because we use the input foreground model to calibrate our visibilities, the gain estimates 
for our two runs differed. In particular we saw the overall flux scale of the calibrated 
visibilities was higher when using the full hierarchical catalog. In order to put both residual 
2D power spectra onto the same scale for comparison, we first divide each pixel by the 
corresponding dirty power spectrum pixel, then subtract to arrive at a ratio difference. 
\begin{equation}\label{eq:diff_ratio}
\mathtt{ratio\;difference = \frac{residual_1}{dirty_1} - \frac{residual_2}{dirty_2}}
\end{equation}
Here we use subscript 1 to represent the MWACS catalog spectra, and 2 for the 
hierarchical catalog spectra. The ratio difference is shown in Figure~\ref{fig:mwacs_ratio}. 
The entire wedge being positive (blue) demonstrates that using the new catalog resulted in 
successfully subtracting higher fractional power. Not only does this indicate we subtract 
more power, but it also confirms that our calibrated data is more closely matched to the 
model, meaning that our calibration solutions in general are more accurate.

@@

We demonstrate the impact of our diffuse model again by running the golden data set 
through our entire analysis pipeline with and without using the diffuse model. Because the 
foreground models used for calibration in the two runs are identical (diffuse is omitted for 
calibration), the calibration solutions were identical and therefore the dirty power spectra 
were identical as well. We compare the model and residual power spectra by direct 
subtraction in Figure~\ref{fig:diffuse_diff}. The top panel shows the model power with the 
diffuse subtracted from the model power without. The entire plot is negative (red) because 
we only added power with the diffuse model. The bottom panel is the difference in residual 
power. The wedge is completely positive (blue), indicating that the diffuse model 
successfully subtracted from the dirty visibilities. The differences in the window are positive 
and negative, indicating that any differences are below the three hour noise level.

@@

>>>>

Using the first generation of instruments, several upper limits have been placed on the 21 
cm EoR signal \citep{Ali:2015, Dillon:2014, Parsons:2014, Jacobs:2015, Paciga:2013}, and 
\citealt{Pober:2015} was able to place constraints on physical reionization models. 
However, much work is to be done to understand the data produced by these arrays. 
Foreground subtraction and isolation has emerged as a priority in the field. Central to most 
analysis strategies is the concept of an ``EoR window" -- the region of Fourier space where 
spectrally smooth foregrounds have been isolated from the isotropic (spherically symmetric 
in Fourier space) cosmological signal \citep{Morales:2006, Bowman:2009}. More recent 
studies have shown the existence of a foreground ``wedge", the result of instrumental mode 
mixing, throwing power from spectrally smooth foregrounds to higher Fourier modes 
\citep{Thyagarajan:2015b, Thyagarajan:2015, Trott:2012, Liu:2014a, Liu:2014b, 
Hazelton:2013, Pober:2013, Vedantham:2012, Morales:2012, Datta:2010}. Despite this 
setback, the EoR window is preserved above the wedge, and the field is pushing forward with 
several analysis pipelines under active development \citep[e.g. B.~J.~Hazelton et al. 2016, in 
prep; D.~A.~Mitchell et al. 2016, in prep; D.~C.~Jacobs et al. 2016, in review; ][]{Trott:2016, 
Dillon:2013, Trott:2014}.

@@

The remainder of this paper is organized as follows. In Section~\ref{sec:MWA} we briefly 
describe the MWA instrument and the observations used in this work, in Section~
\ref{sec:analysis} we describe our analysis pipeline, in Section~\ref{sec:techniques} we 
describe several novel techniques in our analysis, in Section~\ref{sec:deep} we discuss our 
efforts to apply the analysis pipeline to a deep data integration \textbf{and compare with a reference pipeline for robust estimation}, and we discuss future work 
in Section~\ref{sec:discussion}. Throughout this paper we use a $\Lambda$CDM 
cosmology with $\Omega_m=0.73$, $\Omega_\Lambda=0.27$, and $h = 0.7$, consistent 
with WMAP seven year results \citep{Komatsu:2011}. All distances and wavenumbers are in 
comoving coordinates.

@@

The technical design of the MWA is reviewed in \citealt{Tingay:2013}, and we highlight a 
few key characteristics that will become important in our analysis of the data. Each 
antenna of the MWA comprises 16 dual-polarization dipoles placed on a regular grid, lying 
on a ground screen. The radio frequency signals from these dipoles feed into an analog 
beamformer which uses physical delays to ``point" the antenna. The MWA contains 128 of 
these antennas, with a tightly packed core of radius 50 m, and extending out to a radius \textbf{of}
1.5 km for higher resolution calibration and imaging.

@@

\section{Analysis Pipeline}\label{sec:analysis}
In order to ensure consistency in analysis, the international MWA EoR collaboration has 
defined two independent reference analysis pipelines. The details of this strategy are 
outlined in D.~C.~Jacobs et al. 2016 (in review). This study is based on the FHD-to-
\eppsilon pipe, which is in turn based on the Fast Holographic Deconvolution (FHD
\footnote{\url{https://github.com/EoRImaging/FHD}}, \citealt{Sullivan:2012}) and Error 
Propagated Power Spectrum with Interleaved Observed Noise (\eppsilon
\footnote{\url{https://github.com/EoRImaging/eppsilon}}, B.~J.~Hazelton et al. 2016, in 
prep) packages. The general flow of the pipeline is shown in Figure~\ref{fig:pipe}, and we 
give an overview below. \textbf{As a point of comparison, and to demonstrate robustness of the results, we compute the output power spectra with an independent calibration and power spectrum estimation pipeline, and demonstrate consistency.}

@@

Besides RFI, we manually flag data to remove three other effects. As mentioned in 
Section~\ref{sec:MWA}, the edges of the coarse bands contain high levels of aliasing from 
the filter applied in the digital receivers. We flag two 40 kHz channels on either side of the 
coarse band edges (total of four channels per coarse band edge). In principle the aliasing 
could be calibrated out and these channels could be recovered, but this is left for future 
work. Second, early data from the MWA also contained some instances where antennas 
were not pointed properly at the beginning of observations due to a beamformer error. This 
problem has since been resolved, but exists in our data. We therefore flag the first two 
seconds of each observation to avoid any potentially mis-pointed data. Finally, we flag the 
central fine channel of each coarse band because it corresponds to the DC mode of that 
sub-band and can easily float, washing out useful information. \textbf{Please reword this sentence. It is not clear what "float" means here. Can you be more specific?}

@@

\section{Novel Analysis Methods}\label{sec:techniques}
\textcolor{red}{Needs a better section title}
\textbf{Analysis Methods for Testing Data Quality}

The two dimensional power spectrum is a useful figure of merit (FoM) \textbf{(the PS is more of a tool than a FoM)} as we improve and 
refine our analysis pipeline. Foregrounds and systematics often manifest with characteristic 
shapes in this space, enabling us to diagnose problems and quantify improvements 
\citep{Morales:2012}. We use the MWA 3 hour ``golden'' data set from August 23, 2013, to 
repeatedly form power spectra to test and refine our analysis. While an exhaustive catalog 
of these improvements would be unsuited for this article (and likely bore the reader), we 
demonstrate the utility of the 2D power spectrum as a FoM with a few key techniques we 
have employed in our analysis.

@@

Because we use the input foreground model to calibrate our visibilities, the gain estimates 
for our two runs differed. In particular we saw the overall flux scale of the calibrated 
visibilities was higher when using the full hierarchical catalog. In order to put both residual 
2D power spectra onto the same scale for comparison, we first divide each pixel by the 
corresponding dirty power spectrum pixel, then subtract to arrive at a ratio difference. 
\begin{equation}\label{eq:diff_ratio}
\mathtt{ratio\;difference = \frac{residual_1}{dirty_1} - \frac{residual_2}{dirty_2}}
\end{equation}
Here we use subscript 1 to represent the MWACS catalog spectra, and 2 for the 
hierarchical catalog spectra. The ratio difference is shown in Figure~\ref{fig:mwacs_ratio}. 
The entire wedge being positive (blue) demonstrates that using the new catalog resulted in 
successfully subtracting higher fractional power. Not only does this indicate we subtract 
more power, but it also confirms that our calibrated data is more closely matched to the 
model, meaning that our calibration solutions in general are more accurate. \textbf{Consider adding a reference and a sentence about Jack's paper here. Her performed quite a few of these type of tests for understanding the impact of having an incomplete point source model.}

@@

We demonstrate the impact of our diffuse model again by running the golden data set 
through our entire analysis pipeline with and without using the diffuse model. Because the 
foreground models used for calibration in the two runs are identical (diffuse is omitted for 
calibration), the calibration solutions were identical and therefore the dirty power spectra 
were identical as well. We compare the model and residual power spectra by direct 
subtraction in Figure~\ref{fig:diffuse_diff}. The top panel shows the model power with the 
diffuse subtracted from the model power without. The entire plot is negative (red) because 
we only added power with the diffuse model. The bottom panel is the difference in residual 
power. The wedge is completely positive (blue), indicating that the diffuse model 
successfully subtracted from the dirty visibilities. The differences in the window are positive 
and negative, indicating that any differences are below the three hour noise level.   \textbf{Why is the region in the EoR window red here, particularly when the lower plot shows noise-like differences?}

@@

\textbf{The RTS uses a fundamentally different calibration and foreground subtraction approach than FHD, as well as a different input point source model. The primary differences lie in the bandpass calibration, the use of ionospheric corrections, and the lack of a diffuse model. The calibration model uses a model with 1000 sources, for which ionospheric corrections are estimated and applied, before peeling of those sources from the dataset. The bandpass calibration is fundamentally different to FHD... **BART: please also include a mention of the 150m cable reflection and why we may not expect to see it with RTS**. The CHIPS power spectrum estimator applies a full maximum-likelihood estimator to the data. In its full form, the data are weighted by the thermal noise, and a model for the residual point source signal, within a full frequency-dependent description of the instrument. For this comparison, the foreground weighting is not used, in order for a direct comparison with the weighting scheme used by eppsilon.}

\textbf{Of the 1,029 observations calibrated and used by the FHD pipeline, the RTS$+$CHIPS pipeline successfully calibrated and included 1,003. The remainder failed to calibrate successfully, potentially due to adverse ionospheric conditions. The data were divided into the same three redshift bins, and the 2D and 1D power spectra formed. Table \ref{tbl:limits} displays comparison 2-sigma upper limits, demonstrating broad consistency between the pipelines. The RTS$+$CHIPS best limits appear at a lower $k$ bin than FHD$+$eppsilon, owing to the different structure of the foreground leakage for the two pipelines. Figure \ref{fig:slices}, which shows the resulting FHD$+$eppsilon slices in $k_\parallel$ and $k_\bot$, also displays the equivalent power profiles for RTS$+$CHIPS. The comparison pipeline shows significantly more power at low $k_\bot$, consistent with no diffuse model being subtracted from the data. Aside from this area, the two pipelines show consistent results. Figure \ref{fig:ps_comparison} compares the 1D power from both pipelines for the three central redshifts studied in this work. The magenta line reproduces the FHD$+$eppsilon results from Figure \ref{fig:1d_ps}, while the blue line plots the equivalent RTS$+$CHIPS results (with coarse band harmonics included). Despite the differences in the two pipelines, they produce consistent power across a range of scales, supporting the robustness of the results.}
